---
title: "Spells and Spreadsheets: Decoding the Wizarding World with Data Science"
author: "Viviana Lara, Justin Cline"
output: slidy_presentation
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

In the realm of fantasy literature and cinema, few franchises have captured the imagination quite like Harry Potter. But beyond the spellbinding narratives lies a wealth of data waiting to be explored. Our analysis aims to uncover the hidden patterns and insights within the Harry Potter movie series, using a comprehensive dataset that spans characters, dialogue, spells, and movie statistics.

The central question we seek to answer is: How do the various elements of the Harry Potter universe interact and evolve throughout the series? To address this, we'll be diving into multiple CSV files, including:

- **Chapters.csv:** Contains information about movie chapters, including character IDs, chapter names, movie IDs, and chapter numbers within each movie script.
- **Characters.csv:** Provides detailed information about characters, including their species, gender, Hogwarts house, patronus, and wand properties.
- **Dialogue.csv:** Contains every line of dialogue from the movie scripts, linked to characters, chapters, and locations.
- **Movies.csv:** Offers production details for each movie, including title, release year, runtime, budget, and box office performance.
- **Places.csv:** Lists various locations in the Harry Potter universe, categorized by type.
- **Spells.csv:** Details the various spells used in the series, including their incantations, effects, and associated light.

By analyzing this rich dataset, we aim to reveal trends in character development, explore the complexity of magic over time, and even draw connections between the fictional world and real-world factors like movie budgets and audience reception. Whether you're a die-hard fan or a data enthusiast, this analysis promises to shed new light on the intricate tapestry of the wizarding world, all through the lens of data science.



# About the Dataset
```{r message = FALSE, warning = FALSE}
# Loading Required Libraries
library(dplyr)
library(tidyverse) # Needed for a variety of functions related to dataframe manipulation
library(ggplot2) # Needed for plotting data
library(stringr) # Needed for the str_detect() function to manipulate our dataset
library(knitr)      # for kable function
library(kableExtra) # for kable_styling
library(RColorBrewer)
library(tidytext)
library(wordcloud)
```

```{r message = FALSE, warning=FALSE}
# Reading the Datasets
Chapters <- read.csv("HPDatasets/Chapters.csv")
Characters <- read.csv("HPDatasets/Characters.csv")
Data_Dict <- read.csv("HPDatasets/Data_Dictionary.csv")
Movies <- read.csv("HPDatasets/Movies.csv")
Dialogue <- read.csv("HPDatasets/Dialogue.csv")
Places <- read.csv("HPDatasets/Places.csv")
Spells <- read.csv("HPDatasets/Spells.csv")



glimpse(Spells)
glimpse(Places)
glimpse(Dialogue)
```






 

# House Dynamics

This analysis explores the characteristics and demographics of different Hogwarts houses and magical schools using the Characters dataset. We examine several key variables including:

- House affiliation (Gryffindor, Slytherin, Ravenclaw, Hufflepuff, Beauxbatons, Durmstrang)
- Gender distribution within houses
- Blood status diversity (Pure-blood, Half-blood, Muggle-born, etc.)
- Physical characteristics (Hair color, Eye color)
- Magical abilities (Patronus forms)



### Distribution of Characters Across Hogwarts Houses
```{r message = FALSE, warning = FALSE}
Characters %>% # main data frame containing character information
  filter(!is.na(House) & House != "") %>%  # filter out characters without a house and NA
  # mutate creates new columns and transforms the data
  mutate(House_Display = House,  # keeps original names for legend
         House = case_when(
           House == "Beauxbatons Academy of Magic" ~ "Beauxbatons",
           House == "Durmstrang Institute" ~ "Durmstrang",
           TRUE ~ House
         )) %>%
  ggplot(aes(x = House, 
             fill = House_Display)) +
  geom_bar() + # creates bar chart visualization
  geom_text(aes(label = after_stat(count)), 
            position = position_fill(),
            stat = "count", 
            color = "white", 
            size = 3) +
  labs(title = "Distribution of Characters Across Houses",
       x = "House", y = "Count",
       fill = "School") +  # legend title
  theme_minimal() + # minimal theme styling
  # custom colors to houses
  scale_fill_manual(values = c("Gryffindor" = "#740001",
                              "Slytherin" = "#1A472A",
                              "Ravenclaw" = "#0E1A40",
                              "Hufflepuff" = "#FFD800",
                              "Beauxbatons Academy of Magic" = "#87CEEB",
                              "Durmstrang Institute" = "#DC143C")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), # adjust x-axis labels
        legend.text = element_text(size = 8))  # legend text size
```

  


### Gender Distribution
```{r}
Gender.Distribution <- Characters %>%
  filter(!is.na(House) & House != "" & !is.na(Gender)) %>%
  mutate(House = case_when(
    House == "Beauxbatons Academy of Magic" ~ "Beauxbatons",
    House == "Durmstrang Institute" ~ "Durmstrang",
    TRUE ~ House
  )) %>%
  group_by(House, Gender) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(House) %>%
  mutate(Percentage = Count / sum(Count) * 100)

# plot
ggplot(Gender.Distribution, aes(x = House, 
                                y = Percentage, 
                                fill = Gender)) +
  geom_bar(stat = "identity", 
           position = "stack") +
  geom_text(aes(label = sprintf("%.1f%%", Percentage)), 
            position = position_stack(vjust = 0.5), 
            color = "white", size = 3) + # adds percentage inside bars
  labs(title = "Gender Distribution Among Houses",
       x = "House", 
       y = "Percentage") +
  scale_fill_manual(values = c("Female" = "#FF69B4",
                               "Male" = "#4169E1")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



### Blood Status Diversity
```{r message = FALSE, warning = FALSE}
# Filter and prepare data, excluding NA and empty values
Blood.Status <- Characters %>%
  filter(!is.na(House) & House != "" & 
         !is.na(Blood.Status) & Blood.Status != "") %>%
  mutate(House = case_when(
    House == "Beauxbatons Academy of Magic" ~ "Beauxbatons",
    House == "Durmstrang Institute" ~ "Durmstrang",
    TRUE ~ House
  ))

# Calculate percentages
Blood.Status.Percentages <- Blood.Status %>%
  group_by(House, Blood.Status) %>%
  summarise(count = n()) %>%
  group_by(House) %>%
  mutate(percentage = round((count / sum(count)) * 100, 1))

# Create visualization
ggplot(Blood.Status.Percentages, aes(x = House, y = percentage, fill = Blood.Status)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            position = position_stack(vjust = 0.5),
            color = "white", 
            size = 3) +
  labs(title = "Blood Status Distribution Across Houses",
       x = "House",
       y = "Percentage",
       fill = "Blood Status") +
  scale_fill_manual(values = c(
    "Pure-Blood" = "#4A0404",     # Dark red
    "Half-Blood" = "#000080",     # Dark blue
    "Muggle-Born" = "#FFB347",     # Light orange
    "Half-Giant" = "#848482", # Silver
    "Quarter-Veela" = "#87CEEB",
    "Part-Goblin" = "darkgreen"
  )) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_line(color = "#E0E0E0"),
    panel.grid.minor = element_blank()
  )
```


 

### Physical Characteristics
```{r message = FALSE, warning = FALSE}
# Filter and prepare data for hair color analysis
Hair.Color.Distribution <- Characters %>%
  filter(!is.na(House) & House != "" & 
         !is.na(Hair.Color) & Hair.Color != "") %>%
  mutate(House = case_when(
    House == "Beauxbatons Academy of Magic" ~ "Beauxbatons",
    House == "Durmstrang Institute" ~ "Durmstrang",
    TRUE ~ House
  ))

# Calculate percentages
Hair.Color.Percentages <- Hair.Color.Distribution %>%
  group_by(House, Hair.Color) %>%
  summarise(count = n()) %>%
  group_by(House) %>%
  mutate(percentage = round((count / sum(count)) * 100, 1))

# Create visualization
ggplot(Hair.Color.Percentages, aes(x = House, y = percentage, fill = Hair.Color)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            position = position_stack(vjust = 0.5),
            size = 3,
            color = "white") +
  labs(title = "Hair Color Distribution Across Houses",
       x = "House",
       y = "Percentage",
       fill = "Hair Color") +
  scale_fill_manual(values = c(
    "Black" = "#000000",
    "Blonde" = "#E0C092",
    "Brown" = "#72442b",
    "Red" = "orangered",
    "Grey" = "#808080",
    "White" = "#F8EBD5",
    "Bald" = "pink",
    "Silver" = "lightgray",
    "Green" = "lightgreen",
    "Dark" = "#4d2d1a",
    "Auburn" = "#9a3300",
    "Variable" = "purple",
    "Tawny" = "#cca57d"
  )) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_line(color = "#E0E0E0"),
    panel.grid.minor = element_blank()
  )
```

```{r}
# Filter and prepare data for eye color analysis
Eye.Color.Distribution <- Characters %>%
  filter(!is.na(Eye.Color) & Eye.Color != "") %>%
  group_by(Eye.Color) %>%
  summarise(
    Count = n(),
    Percentage = round(n() / nrow(filter(Characters, !is.na(Eye.Color) & Eye.Color != "")) * 100, 1)
  ) %>%
  arrange(desc(Percentage))

# Create visualization
ggplot(Eye.Color.Distribution, aes(x = reorder(Eye.Color, Percentage), y = Percentage, fill = Eye.Color)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", Percentage)), 
            hjust = -0.1) +
  coord_flip() +
  labs(title = "Distribution of Eye Colors",
       x = "Eye Color",
       y = "Percentage") +
  scale_fill_manual(values = c(
    "Blue" = "#1E90FF",
    "Brown" = "#8B4513",
    "Green" = "#228B22",
    "Grey" = "#808080",
    "Black" = "#000000",
    "Dark" = "#2F4F4F",
    "Hazel" = "#DEB887",
    "Yellow" = "yellow",
    "Silver" = "lightgrey",
    "Scarlet" = "#FF2400",
    "Pale" = "beige",
    "Ruddy" = "#ff0028",
    "Variable" = "purple",
    "Gooseberry" = "#E6EAC0"
  )) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank()
  ) +
  scale_y_continuous(limits = c(0, max(Eye.Color.Distribution$Percentage) * 1.2))
```

 

### Patronus Forms
```{r}
# Count frequency of each Patronus form
Patronus.Count <- Characters %>%
  filter(!is.na(Patronus) & Patronus != "") %>%
  group_by(Patronus) %>%
  summarise(
    Count = n(),
    Percentage = round(n() / sum(!is.na(Characters$Patronus) & Characters$Patronus != "") * 100, 1)
  ) %>%
  arrange(desc(Count))

# Display table using kable
kable(head(Patronus.Count, 10),
      col.names = c("Patronus Form", "Number of Characters", "Percentage"),
      caption = "Top 10 Most Common Patronus Forms") %>%
  kable_styling(bootstrap_options = "striped")
```

```{r warning=FALSE}
# Join Characters table to get house and patronus information
House.Patronus <- Characters %>%
  filter(!is.na(House) & House != "" & !is.na(Patronus) & Patronus != "") %>%
  mutate(House = case_when(
    House == "Beauxbatons Academy of Magic" ~ "Beauxbatons",
    House == "Durmstrang Institute" ~ "Durmstrang",
    TRUE ~ House
  ))

# Create visualization
ggplot(House.Patronus, aes(x = House, fill = Patronus)) +
  geom_bar(position = "fill") +
  labs(title = "Distribution of Patronus Forms by House",
       x = "House",
       y = "Proportion") +
  scale_fill_viridis_d() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


 

# Dialogue Analysis
To analyze dialogue patterns across the Harry Potter movie series, we examined the following datasets:

- Dialogue.csv: Contains character dialogue lines with Character.ID and Chapter.ID
- Characters.csv: Provides character details including House affiliation
- Chapters.csv: Links dialogues to specific movie chapters
- Movies.csv: Contains movie titles and chronological information

Key variables analyzed include:

- Individual dialogue lines and their frequency
- Distribution of dialogue across Houses
- Speaking patterns of main characters (Golden Trio)
- Common words and phrases used throughout the series

 

### Frequently Used Words

```{r message=FALSE, warning=FALSE}


# Prepare the data
word_counts <- Dialogue %>%
  unnest_tokens(word, Dialogue) %>%
  anti_join(stop_words) %>%
  count(word, sort = TRUE)

# Create the word cloud
set.seed(1234)
wordcloud(words = word_counts$word, 
          freq = word_counts$n, 
          min.freq = 1,
          max.words = 100, 
          random.order = FALSE, 
          rot.per = 0.35, 
          colors = brewer.pal(8, "Dark2"))
```


```{r message=FALSE, warning=FALSE}
# Create frequency table of most common words
Word.Freq.Table <- Dialogue %>%
  unnest_tokens(word, Dialogue) %>%
  anti_join(stop_words) %>%
  count(word, sort = TRUE) %>%
  top_n(20) %>%
  mutate(percentage = n/sum(n)*100)

# Display table using kable
kable(Word.Freq.Table,
      col.names = c("Word", "Frequency", "Percentage"),
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  row_spec(0, bold = TRUE) %>%
  add_header_above(c(" " = 1, "Word Statistics" = 2))
```



 

### Locations with Most Dialogue
```{r}
# Count dialogues per location
Dialogue.Count <- Dialogue %>%
  group_by(Place.ID) %>%
  summarise(Dialogue_Count = n()) %>%
  arrange(desc(Dialogue_Count))

# Join with places to get place names
TopLocations <- Dialogue.Count %>%
  left_join(Places, by = "Place.ID") %>%
  top_n(10, Dialogue_Count)

# Visualize top 10 locations with most dialogue
ggplot(TopLocations, aes(x = reorder(Place.Name, Dialogue_Count), y = Dialogue_Count)) +
  geom_bar(stat = "identity",
           fill = "#2A623D") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 10 Locations with Most Dialogue",
       x = "Location",
       y = "Number of Dialogue Lines") +
  theme(axis.text.y = element_text(angle = 0, hjust = 1))
```


 

### Representation of Different Houses in Dialogue Across the Movie Series
```{r}
# Join Characters and Dialogue tables to get house information for each dialogue
House.Dialogue <- Dialogue %>%
  left_join(Characters, by = "Character.ID") %>%
  filter(!is.na(House) & House != "") %>%
  # Simplify school names
  mutate(House = case_when(
    House == "Beauxbatons Academy of Magic" ~ "Beauxbatons",
    House == "Durmstrang Institute" ~ "Durmstrang",
    TRUE ~ House
  ))

# Count dialogues per house per movie
House.Representation <- House.Dialogue %>%
  left_join(Chapters, by = "Chapter.ID") %>%
  left_join(Movies, by = "Movie.ID") %>%
  group_by(Movie.Title, House) %>%
  summarise(Dialogue_Count = n(), .groups = 'drop') %>%
  group_by(Movie.Title) %>%
  mutate(Percentage = Dialogue_Count / sum(Dialogue_Count) * 100,
         # Create shorter movie names
         Movie.Short = case_when(
           Movie.Title == "Harry Potter and the Philosopher's Stone" ~ "HP 1",
           Movie.Title == "Harry Potter and the Chamber of Secrets" ~ "HP 2",
           Movie.Title == "Harry Potter and the Prisoner of Azkaban" ~ "HP 3",
           Movie.Title == "Harry Potter and the Goblet of Fire" ~ "HP 4",
           Movie.Title == "Harry Potter and the Order of the Phoenix" ~ "HP 5",
           Movie.Title == "Harry Potter and the Half-Blood Prince" ~ "HP 6",
           Movie.Title == "Harry Potter and the Deathly Hallows Part 1" ~ "HP 7.1",
           Movie.Title == "Harry Potter and the Deathly Hallows Part 2" ~ "HP 7.2"
         ))

# Create visualization
ggplot(House.Representation, 
       aes(x = Movie.Short, 
           y = Percentage, 
           fill = House)) +
  geom_bar(stat = "identity", 
           position = "stack") +
  labs(title = "House Representation in Dialogue Across Movies",
       x = "Movie",
       y = "Percentage of Dialogue") +
  scale_fill_manual(values = c(
    "Gryffindor" = "#740001",
    "Slytherin" = "#1A472A",
    "Ravenclaw" = "#0E1A40",
    "Hufflepuff" = "#FFD800",
    "Beauxbatons" = "#87CEEB",
    "Durmstrang" = "#DC143C"
  )) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
# Create summary table of house dialogue representation
House.Summary <- House.Dialogue %>%
  left_join(Chapters, by = "Chapter.ID") %>%
  group_by(House) %>%
  summarise(
    Total.Lines = n(),
    Percentage = round(n() / nrow(House.Dialogue) * 100, 1)
  ) %>%
  arrange(desc(Total.Lines))

# Display table using kable
kable(House.Summary,
      col.names = c("House", "Total Lines of Dialogue", "Percentage of All Dialogue"),
      caption = "Overall House Representation in Dialogue") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


 

### Golden Trio Dialogue Distribution
```{r warning=FALSE, message=FALSE}
# Analyze dialogue distribution for main characters across movies
main_char_dialogue <- Dialogue %>%
  left_join(Characters, by = "Character.ID") %>%
  left_join(Chapters, by = "Chapter.ID") %>%
  left_join(Movies, by = "Movie.ID") %>%
  filter(Character.Name %in% c("Harry Potter", "Ron Weasley", "Hermione Granger")) %>%
  group_by(Movie.Title, Character.Name) %>%
  summarise(Dialogue_Count = n()) %>%
  mutate(Movie.Short = case_when(
    Movie.Title == "Harry Potter and the Philosopher's Stone" ~ "HP 1",
    Movie.Title == "Harry Potter and the Chamber of Secrets" ~ "HP 2",
    Movie.Title == "Harry Potter and the Prisoner of Azkaban" ~ "HP 3",
    Movie.Title == "Harry Potter and the Goblet of Fire" ~ "HP 4",
    Movie.Title == "Harry Potter and the Order of the Phoenix" ~ "HP 5",
    Movie.Title == "Harry Potter and the Half-Blood Prince" ~ "HP 6",
    Movie.Title == "Harry Potter and the Deathly Hallows Part 1" ~ "HP 7.1",
    Movie.Title == "Harry Potter and the Deathly Hallows Part 2" ~ "HP 7.2"
  ))

# Create visualization
ggplot(main_char_dialogue, 
       aes(x = Movie.Short, 
           y = Dialogue_Count, 
           color = Character.Name,
           group = Character.Name)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Golden Trio Dialogue Distribution Across Movies",
       x = "Movie",
       y = "Number of Dialogue Lines") +
  scale_color_manual(values = c(
    "Harry Potter" = "#740001",
    "Ron Weasley" = "#CD7F32",
    "Hermione Granger" = "#964B00"
  )) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_blank()
  )
```



 

# Spell Analysis
To find spell usage by each character and house we utilize a function 
to search through every line of dialogue searching for occurrences of spell 
incantations and then record the which character used the spell and what spell
they used.

To analyze spell usage throughout the Harry Potter movies we examined the following datasets:

- Dialogue.csv: Contains each line of dialogue and who said it

- Spells.csv: Contains the spell incantations

Key Variables include:

- Individual dialogue lines and the incantations they contain

- Character names 

- House names

- Incantations (e.g. Sectum Sempra, Expelliarmus)

### Spell Finder Function
```{r, echo=TRUE}
SpellFinder <- function(dialogueDF, spellsDF) {
  #Initializes up the matches data frame to hold the 
  #line of dialogue and spell being used in that line of dialogue.
  Matches <- data.frame(
    Dialogue = character(),
    Spell = character()
  )
  #for each spell in the Spells data frame we identify which lines of 
  #dialogue contain that spell and add those lines of dialogue to the MatchingLines vector.
  for (spell in spellsDF$Incantation) {
    MatchingLines <- dialogueDF$Dialogue[str_detect(
      dialogueDF$Dialogue,spell)]
      
#Stores the vector MatchingLines in a temporary data 
#frame and Labels each dialogue with the spell found in them from this iteration of the loop.
    MatchesTemp <- data.frame(
      Dialogue = MatchingLines,
      Spell = rep(spell, length(MatchingLines))
    )
#Once we have a temporary data frame with each of the 
#lines of dialogue and have labeled them by the 
#spell used we can add the dialogues with this 
#iterations spells to a final data frame which will hold 
#all lines of dialogue with a spell from all iterations at the end of the loop
      Matches <- rbind(Matches,MatchesTemp)
      }
  
  return(Matches)
}

```

### Spell Usage by the golden trio
```{r message = FALSE, warning=FALSE}
SpellUsage<-inner_join(SpellFinder(Dialogue, Spells),Dialogue) %>% 
  arrange(Dialogue.ID)


SpellsByCharacter<-inner_join(SpellUsage,Characters)



SpellCountByCharacter<-SpellsByCharacter %>% 
  group_by(Character.Name, Character.ID, Spell,Chapter.ID) %>% 
  summarise("NumberofCasts"=n())



SpellCountByCharacter %>% 
filter(Character.ID == 1 | Character.ID == 2 | Character.ID == 3 ) %>% 
  ggplot(aes(x=Character.Name, y=NumberofCasts))+
  geom_col()+
  facet_wrap('Spell')+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )+
  labs(title = "Number of Times the Golden Trio Cast Each Spell", y = "Number of Casts", x= "Character")
```


### How many times each spell has been used as the movies progress
```{r, message = FALSE, warning=FALSE}
SpellCountByChapter<-SpellCountByCharacter %>% 
  group_by(Spell,Chapter.ID) %>% 
  summarise('CastsPerChapter'=n())%>%
  mutate(CumulativeCasts = cumsum(CastsPerChapter)) %>% 
  filter( n_distinct(Chapter.ID)>1)

SpellCountByChapter %>% 
  ggplot(aes(x=Chapter.ID,y=CumulativeCasts))+
  geom_line()+
  facet_wrap('Spell')+
  labs(title = "Number of Times Each Spell Was Cast", y= "Number of Casts", x= "Chapter")



```

### Which spells are used most frequently by each character
```{r message = FALSE, warning=FALSE}
SpellCountByCharacter %>% 
  ggplot(aes(x=Character.Name, y=Spell, fill =NumberofCasts))+
  geom_tile()+
  scale_fill_gradient(low = "orange", high = "blue")+
  theme(axis.text.y = element_text(size = 6,) ,axis.text.x = element_text(angle = 45, hjust= 1))+
  labs(title = "Spell Usage Heatmap",x = "Character", y= "Number of Casts")

```

### Which spells are used most frequently by each house
```{r message = FALSE, warning=FALSE}
SpellCountByHouse<-SpellsByCharacter %>% 
  group_by(House, Spell) %>% 
  summarise("NumberofCasts"=n()) %>% 
  mutate(CumulativeCasts = cumsum(NumberofCasts)) %>% 
  group_by(House) %>% 
  filter(!is.na(House) & House != "")




SpellCountByHouse %>% 
ggplot(aes(x=House,y=Spell, fill = NumberofCasts))+
  geom_tile(height=1)+
  scale_fill_gradient(low = "orange", high = "blue")+
  theme(axis.text.y = element_text(size = 8))+
  labs(title = "Spell Usage Heatmap")


  
```

### Which spells are most frequently used by each gender
```{r message = FALSE, warning=FALSE}
SpellCountByGender<-SpellsByCharacter %>% 
  group_by(Gender, Spell) %>% 
  summarise("NumberofCasts"=n())




SpellCountByGender  %>% 
ggplot(aes(x=Gender,y=Spell, fill = NumberofCasts))+
  geom_tile(height=1)+
  scale_fill_gradient(low = "orange", high = "blue")+
  theme(axis.text.y = element_text(size = 8))+
  labs(title = "Spell Usage Heatmap")


  
```



# Wand Analysis
To analyze wand characteristics and their distribution across the Harry Potter universe, we examined the following datasets:

- Characters.csv: Contains wand details including wood type, core, and length

Key variables analyzed include:

- Wand length (measured in inches)
- Wand wood type (e.g., Holly, Oak, Vine)
- Wand core material (e.g., Phoenix feather, Dragon heartstring)
- Character house affiliation

 

### Distribution of Wands Across Houses
```{r}
# Create wand count analysis by house
Wand.House.Count <- Characters %>%
  filter(!is.na(House) & House != "" & !is.na(Wand..Wood.) & Wand..Wood. != "") %>%
  mutate(House = case_when(
    House == "Beauxbatons Academy of Magic" ~ "Beauxbatons",
    House == "Durmstrang Institute" ~ "Durmstrang",
    TRUE ~ House
  )) %>%
  group_by(House) %>%
  summarise(
    Wand_Count = n(),
    Percentage = round(n() / nrow(Characters) * 100, 1)
  ) %>%
  arrange(desc(Wand_Count))

# Create bar plot with adjusted text position
ggplot(Wand.House.Count, aes(x = reorder(House, -Wand_Count), y = Wand_Count, fill = House)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(Wand_Count, "\n(", Percentage, "%)"), 
            y = Wand_Count + max(Wand_Count) * 0.05), # Position text above bars
            position = position_dodge(width = 0.9),
            vjust = 0) +
  labs(title = "Distribution of Wands Across Houses",
       x = "House",
       y = "Number of Wands") +
  scale_fill_manual(values = c(
    "Gryffindor" = "#740001",
    "Slytherin" = "#1A472A",
    "Ravenclaw" = "#0E1A40",
    "Hufflepuff" = "#FFD800",
    "Beauxbatons" = "#87CEEB",
    "Durmstrang" = "#DC143C"
  )) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  # Expand y-axis to make room for labels
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)))
```

 

### Wand Length Distribution by Wood Type and Core
```{r}
# Create faceted scatter plot for wand characteristics
Wand.Analysis <- Characters %>%
  filter(!is.na(Wand..Length.) & !is.na(Wand..Core.) & !is.na(Wand..Wood.) & 
         Wand..Core. != "" & Wand..Wood. != "") %>%
  mutate(Wand..Core. = factor(Wand..Core.),
         Wand..Wood. = factor(Wand..Wood.))

# Create the visualization
ggplot(Wand.Analysis, aes(x = Wand..Wood., y = Wand..Length., color = Wand..Core.)) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.6) +
  labs(title = "Wand Length Distribution by Wood Type and Core",
       x = "Wood Type",
       y = "Length (inches)",
       color = "Wand Core") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 8),
    legend.position = "bottom"
  ) +
  scale_color_brewer(palette = "Set2")
```


### Summary

- We found many interesting pieces of information from our exploration of the data, for instance, Ron uses almost no magic throughout the movies,
not a single Hufflepuff casts throughout the movies, and by percentage there are more half-bloods in Slytherin than any other house.

- We also found some less surprising information through our explorations such as Harry is the most common word said throughout the series, Expelliarmus 



